Title

### 2015-07-26 main.tex
Structure-of-data section, and Bonnie edits

This growth in data poses significant challenges for 
researchers~\cite{marx2013biology}.
Currently, many omics applications require us to store, access, and analyze
large libraries of data.
One approach to solving these challenges is to embrace cloud computing.
Google, Inc. and the Broad Institute have teamed up to bring the GATK (Genome 
Analysis Toolkit) to the Google cloud (https://cloud.google.com/genomics/gatk).
Amazon Web Services are also commonly used for computational biology 
research~\cite{schatz2010cloud}.
However, while cloud computing frees researchers from maintaining their own
data centers, and provides cost-saving benefits when computing resources are
not needed around the clock, it is no panacea.
First, in the face of disease outbreaks such as the 2014 Ebola virus epidemic 
in West Africa, analysis resources are needed at often-remote field sites.
While it is now possible to bring sequencing equipment and limited computing 
resources to remote sites, internet connectivity is still highly constrained.
Thus, accessing cloud resources for analytics may not be possible.
Moreover, cloud computing does not truly address the problem posed by
a corpus of data that is growing faster than Moore's law, because the computer
systems that make up those cloud datacenters are still themselves bound by
improvements in semiconductor technology; they still follow the trajectory
suggested by Moore's law.

Computer scientists routinely exploit peculiarities in the structure of data in
order to reduce time or space complexity.
In computational biology, this approach has served researchers well.
Now-classical approaches such as principal component analysis (PCA) reduce the 
dimensionality of data in order to simplify analysis and uncover salient 
features~\cite{berger2013computational}.
As another example, clever indexing techniques such as the Burroughs-Wheeler 
Transform (BWT) also take advantage of aspects of sequence 
data~\cite{berger2013computational}.
This review focuses on algorithmic advances for dealing with the growth in 
biological data; algorithms for gaining novel biological insights are not its 
focus.

@@

Most but not all biological data take the form of \emph{sequence} data, either 
nucleotide sequences (nominally a four-letter alphabet representing the four 
DNA or RNA bases) or protein sequences (nominally a twenty-letter alphabet 
representing the twenty standard amino acids).
Sequence data are obtained in several ways, including mass spectrometry (which
can determine protein sequence) and RNA-seq (which can determine RNA sequence, 
and by inference the protein sequence into which it might translate).
However, the greatest volume of sequence data derive from next-generation 
sequencing (commonly abbreviated NGS) technologies, and comprise DNA sequences.
At the dawn of the genomic era, Sanger sequencing was the most widely used 
method.
Slow and labor intensive but well-tested, Sanger sequencing is still used to 
validate NGS results.
More recently, Sanger sequencing has largely been supplanted by NGS approaches,
beginning with Illumina's ``sequencing by synthesis,'' which allows vastly 
greater throughput due to massive parallelism, low cost, and simple preparation
of samples.
Importantly to this Review, Illumina sequencing, as well as other approaches
such as SOLiD, Ion Torrent, and 454 pyrosequencing, does not read a single DNA
molecule end-to-end as one could a file from a hard drive.
Instead, DNA molecules are chopped into many small fragments, or 
\emph{reads} (Figure ~\ref{fig:shotgun}), 
which are read concurrently, and which must be reassembled in the correct order 
if an entire genome is to be read.
Reads most typically range from 50 to 200 bases long, though longer reads are
available with some technologies.
This so-called ``shotgun sequencing'' is analogous to the problem of 
reconstructing a large file from small, 
randomly-ordered reads, or a shredded paper document.
However, these NGS reads overlap, and they are many in number, so 
reconstructing an entire genome is feasible.
No sequencing technology is completely infallible, so sequencing machines also
provide a \emph{quality score} (or measure of the confidence in the DNA base
called) associated with each position.
Thus, an NGS read is a string of DNA letters, coupled with a string of ASCII
characters that encode the quality of the base call.
A sequencing run will produce many of these reads, typically stored in a FASTQ
file.

@@

Chemical compounds represent another source of structural data, more general
than proteins.
Determined by X-ray crystallography, NMR, electron microscopy and other 
techniques, compounds---many of which are useful pharmaceuticals---can be
represented as labeled graphs.
In such a graph, the vertices are atoms (labeled with chemical element) and
the edges are bonds between atoms (labeled with bond type, such as single or
double).
Chemical compounds are stored in databases such as PubChem, and can be described
as adjacency matrices.

\section{Challenges in biological data}

Given DNA or RNA reads from NGS technologies, the first task is to 
\emph{assemble} those fragments of sequence (which have associated quality 
scores) into contiguous sequences.
The task of \emph{de novo} assembly is beyond the scope of this Review, but is
possible because reads overlap and because the sequence is covered by many such
overlapping reads~\cite{zhang2011practical}; the \emph{de Bruijn graph} is a 
commonly-used data structure.
Often, however, a \emph{reference genome} (or in the case of RNA, 
\emph{transcriptome}) is available for the organism being sequenced; the
establishment of a human reference genome was indeed the purpose of the Human
Genome Project.
When a reference sequence is available, NGS reads must be \emph{mapped} onto
this reference (Figure ~\ref{fig:ngs-pipeline}).
This allows the differences between the newly-sequenced genome and the reference
to be analyzed; these differences may include single-nucleotide polymorphisms
(SNPs, which are the genetic analogue to bit-flips) (Figure~\ref{fig:snps}), 
insertions 
or deletions, or larger-scale changes in the genome.
While reference-based read mapping is a fundamentally simpler problem than
\emph{de novo} assembly, it is still computationally complex, as gigabytes or
terabytes of reads must each be mapped onto the reference genome, which can 
range from millions (for bacteria) to billions (for mammals).

Standard techniques for reference-based read mapping rely on algorithmic 
approaches such as the Burrows-Wheeler transform (BWT)~\cite{burrows1994block}, 
which provides efficient string 
compression through a reversible transformation, while the FM-index data 
structure~\cite{ferragina2000opportunistic} is a compressed substring index, 
based on the BWT, which provides efficient storage as well as fast search.
BWA (Burrows-Wheeler Aligner)~\cite{li2009fast} uses the BWT, while the 
Bowtie2~\cite{langmead2012fast} aligner further relies on the FM-index for 
efficient mapping of NGS reads.
The Genome Multitool (GEM) mapper~\cite{marco2012gem} also uses an FM-index 
coupled with dynamic programming in a compressed representation, in order to 
prune the search space when mapping reads to a reference genome.
Masai~\cite{siragusa2013fast} uses an ``approximate seed'' approach to indexing
the space of possible matches, likewise pruning the search space.
Even with these approaches, read mapping remains a significant bottleneck in
genomic research~\cite{hach2010mrsfast}.

@@

Given a sequenced genome, the next natural questions are what genes (genomic
regions that code for proteins) are present, what structure each resulting
protein takes, and what biological function it performs.
Identifying likely genes is a well-studied problem~\cite{burge1998finding} 
beyond the scope of this Review.
However, determining evolutionary relationships, structure, and function are
at the heart of current research in computational biology.
Since some organisms (known as \emph{model organisms}) are better-studied than
others, and evolution is known to conserve sequence, structure, and function,
a powerful approach to determine these attributes is to search for similar
sequences about which more is known.
This so-called \emph{homology search} entails searching for \emph{approximate}
matches in databases of known gene or protein sequences.

Basic Local Alignment Search Tool (BLAST)~\cite{altschul1990basic} has been
the standard too for performing homology search on databases of nucleotide and
protein sequences.
BLAST takes a ``seed-and-extend'' approach; it looks for small, $k$-mer matches
that might lead to longer matches, and greedily extends them, ultimately 
producing a \emph{sequence alignment} between a query and each potential 
database hit.
However, BLAST's running time is linear in the size of the database being
searched, which is problematic as sequence databases continue to grow at a
faster rate than Moore's law.

Driven by the plummeting costs of next-generation sequencing, the 1000 Genomes Project~\cite{10002012integrated} is pursuing a broad catalog of human 
variation; instead of producing a single reference genome for a species, many
complete genomes are catalogued.
Likewise, WormBase~\cite{harris2014wormbase} and 
FlyBase~\cite{tweedie2009flybase} are cataloguing many different species and
strains of the \emph{Caenorhabditis} worm and \emph{Drosophila} fruit fly, 
respectively.
On a potentially even larger scale is the growth of metagenomic data.
Metagenomics is the study of the many genomes (bacterial, fungal, and even 
viral) that make up a particular environment.
Such an environment could be soil from a particular region (which can lead to 
the discovery of new 
antibiotics~\cite{gillespie2002isolation,forsberg2012shared}), or it could be 
the human gut, whose microbiome has been linked to human-health concerns 
including Autism Spectrum Disorder~\cite{macfabe2012short}, 
Crohn's Disease~\cite{gevers2014treatment,manichanh2006reduced}, and 
obesity~\cite{greenblum2012metagenomic}.

Metagenomics fundamentally asks what organisms are present, and (in the case
of a microbiome such as the gut) what metabolic functions it can accomplish as
a whole.
One approach to this is to attempt to map NGS reads from a metagenomic sample
onto a set of reference genomes that are expected to be present.
This is exactly the read-mapping problem discussed above, but with many 
reference genomes, compounding the computational requirements.
A second approach is to perform homology search on a protein sequence database;
exact or nearly-exact matches imply the presence of a species, while more 
distant hits may still give clues to function.
For this task, BLASTX~\cite{altschul1990basic} is commonly
used to translate nucleotide reads into their possible protein sequences, and
search for them in a protein database.
Recent advances in metagenomic search tools have relied on two advances over
BLASTX: indexing and alphabet reduction.
RapSearch2~\cite{zhao2012rapsearch2} relies on alphabet reduction and a 
collision-free hash table.
The alphabet reduction, as it is reversible, can be thought of as a form of
lossless compression; a 20-letter amino acid alphabet is mapped onto a smaller
alphabet, with offsets stored to recover the original sequence in the full
alphabet.
The hash-table provides an efficient index of the database to be searched.
DIAMOND~\cite{buchfink2014fast} also relies on alphabet reduction, but uses
``shaped seeds'' instead of simple $k$-mer seeds to index the database.
DIAMOND demonstrates search performance three to four orders of magnitude faster
than BLASTX, but it is still linear in the size of the database being searched.

@@

Another important type of biological data is gene expression, or the quantity of
a particular gene product (usually a protein) present in the cell.
Gene expression data is useful for relating genotype to phenotype, and is 
important in studying diseases including cancer.
Usually obtained through microarrays or RNA-seq technologies, expression data is
quantitative, as each gene from a sample is associated with a numeric expression
level, and high-dimensional, as many thousands of genes may be analyzed at a
time.
While expression data lends itself to cluster analysis and probabilistic 
approaches, the high dimensionality presents challenges~\cite{nguyen2002tumor}.
Principal Component Analysis has shown promise in reducing the dimensionality
of gene expression data~\cite{raychaudhuri2000principal,alter2000singular,yeung2001principal,misra2002interactive}.
Some recent work has explored other ways to exploit the high-dimensional 
structure of the data.
SPARCLE (SPArse ReCovery of Linear combinations of Expression)~\cite{prat2011recovering}
brings ideas from compressed sensing~\cite{candes2005decoding,candes2006stable,donoho2006compressed} 
to gene expression analysis.
Specifically, SPARCLE relies on the idea that a randomly-chosen set of points 
are unlikely to be close in a lower-dimensional linear subspace.
Thus, SPARCLE approximates a gene's expression profile as a linear combination
of a few other genes' profiles, and then applies supervised learning to
predict relationships between genes.

Another recent and novel approach to exploiting the structure of gene expression
space is Parti (Pareto task inference)~\cite{hart2015inferring}, which describes a set of
data as a polytope, and infers the specific tasks represented by vertices of
that polytope from the features most highly enriched at those vertices.
In human breast tumors, and in mouse tissues, the expression data were
well-described by tetrahedra whose vertices were enriched for different tumor
types and biological functions, from which~\cite{hart2015inferring} infer four distinct
tasks.

Networks are frequently used to represent biological data, such as the genetic
and physical interactions among proteins, such as those in metabolic pathways.
Several approaches have taken advantage of the particular topologies of these
networks, such as diffusion-based approaches, which explore the topology of
networks through random walks.
IsoRank~\cite{singh2008global} and IsoRankN~\cite{liao2009isorankn} perform global multiple network 
alignment based on network topology and conserved biological function.
IsoRank's approach is similar to the PageRank algorithm~\cite{page1999pagerank}, while
IsoRankN produces a \emph{Personalized PageRank} alignment, similar to the
PageRank-Nibble algorithm~\cite{andersen2006local}.
In protein interaction networks, some nodes are high-degree \emph{hubs,} which 
interact with
many partners; these interactions do not imply shared function as strongly as
non-hub interactions.
Diffusion state distance (DSD)~\cite{cao2013going} is another diffusion-based approach
which defines a distance metric based on random walks.
DSD penalizes high-degree nodes, and outperforms other graph-theoretic distance
functions when applied to standard approaches such as neighbor voting for function annotation.

As new functional annotations and protein-protein interactions are discovered,
the number and size of these networks grows.
Diffusion component analysis (DCA)~\cite{cho2015diffusion} amounts to a network-based 
analogue of principle component analysis; DCA reduces the dimensionality of a
network in such a way that function annotation can be performed in the reduced 
space, significantly reducing running time while still providing competitive
prediction accuracy.

@@

@@

One key insight as to why the ``compressive omics'' approach allows computation
that scales sublinearly in the size of the data is that such data sets exhibit
low \emph{metric entropy}~\cite{yu2015entropy}.
That is, for a given cluster radius $r_c$ and a database $D$, the number 
$N_{r_c} (D)$ of clusters needed to cover $D$ is relatively small compared to
$|D|$, the number of entries in the database.
In contrast, if the points were uniformly distributed about the Cartesian space,
$N_{r_c} (D)$ would be larger.

A second key insight is that the biological data sets considered 
by~\cite{yu2015entropy} exhibit low \emph{fractal dimension}.
That is, for some range of radii $r_1$ and $r_2$,

discrete?

By combining redundancy with a low fractal 
dimension~\cite{yu2015entropy}, it may be possible to develop algorithms that 
scale even as the data continue to grow.

wherever physical or empirical laws have constrained data to 
a subspace of low entropy and fractal dimension.

Specifically, the ratio $\frac{|D|}{k}$ provides an estimate of the acceleration factor
for just the coarse search component compared to a full linear search.
Local fractal dimension around a data point can be computed by determining the
number of other data points within two radii $r_1$ and $r_2$ of that point;
given those point counts ($n_1$ and $n_2$, respectively), fractal dimension $d$
is simply $d=\frac{\log (n_2 / n_1)}{ \log (r_2 / r_1)}$.
Sampling this property over a dataset can provide a global average fractal 
dimension.
When we search a larger radius around a query, the number of points we encounter grows exponentially with the fractal dimension;
low fractal dimension implies that this growth will not obviate the gains provided by an entropy-scaling data structure.

An entropy-scaling search tree is also related to a metric ball tree~\cite{uhlmann1991satisfying}, although with different time complexity.
Querying a metric ball tree requires $O(\log n)$ time, assuming the relatively uniform distribution of data points in a metric space.
This is different from the nonuniform distribution under which entropy-scaling
search behaves well.

\begin{table}
\caption{Metric entropy ratio (ratio of clusters to entries in database) and
fractal dimension at typical search radii for four biological data sets.
NCBI's ``NR'' non-redundant protein sequence and ``NT'' non-redundant nucleotide sequence databases are from June, 2015. Protein Databank (PDB) is from July, 2015. PubChem is from October, 2013.}
\begin{tabular}{ccc}
\hline
Data set & Metric entropy ratio & Fractal dimension \\        
\hline
Nucleotide sequences (NCBI NT) & 7:1 & 1.5\\
\hline
Protein sequences (NCBI NR) & 5:1 & 1.6\\
\hline
Protein structure (PDB) & 10:1 & 2.5\\
\hline
Chemical structure (PubChem) & 11:1 & 0.2\\
\hline
\end{tabular}
\end{table}

>>>>

This growth in data poses significant challenges for 
researchers~\cite{marx2013biology}.
Currently, many biological 'omics` applications require us to store, access, and analyze
large libraries of data.
One approach to solving these challenges is to embrace cloud computing.
Google, Inc. and the Broad Institute have teamed up to bring the GATK (Genome 
Analysis Toolkit) to the Google cloud (https://cloud.google.com/genomics/gatk).
Amazon Web Services are also commonly used for computational biology 
research and enterprise (e.g., DNANexus)~\cite{schatz2010cloud}.
However, while cloud computing frees researchers from maintaining their own
data centers, and provides cost-saving benefits when computing resources are
not needed around the clock, it is no panacea.
First and foremost, cloud computing does not truly address the problem posed by
a corpus of data that is growing faster than Moore's law, because the computer
systems that make up those cloud datacenters are still themselves bound by
improvements in semiconductor technology; they still follow the trajectory
suggested by Moore's law.
Moreover, in the face of disease outbreaks such as the 2014 Ebola virus epidemic 
in West Africa, analysis resources are needed at often-remote field sites.
While it is now possible to bring sequencing equipment and limited computing 
resources to remote sites, internet connectivity is still highly constrained.
Thus, accessing cloud resources for analytics may not be possible.

Computer scientists routinely exploit the structure of various data in
order to reduce time or space complexity.
In computational biology, this approach has implicitly served researchers well.
Now-classical approaches such as principal component analysis (PCA) reduce the 
dimensionality of data in order to simplify analysis and uncover salient 
features~\cite{berger2013computational}.
As another example, clever indexing techniques such as the Burroughs-Wheeler 
Transform (BWT) also take advantage of aspects of sequence 
data~\cite{berger2013computational} to speed up computation and save storage.
This review focuses on cutting-edge algorithmic advances for dealing with the growth in 
biological data by explicitly taking advantage of its unique structure; algorithms for gaining novel biological insights are not its 
focus.

@@

Most but not all biological data take the form of \emph{sequence} data, either 
nucleotide sequences (nominally a four-letter alphabet representing the four 
DNA or RNA bases) or protein sequences (nominally a twenty-letter alphabet 
representing the twenty standard amino acids).
Sequence data are obtained in several ways, including mass spectrometry (which
can determine protein sequence, as well as interactions) and RNA-seq (which can determine RNA sequence, 
and by inference the expression  of the gene it might translate).
However, the greatest volume of sequence data derive from next-generation 
sequencing (commonly abbreviated NGS) technologies, and comprise DNA sequences.
At the dawn of the genomic era, Sanger sequencing was the most widely-used 
method.
Slow and labor intensive but well-tested, Sanger sequencing is still used to 
validate NGS results.
More recently, Sanger sequencing has largely been supplanted by NGS approaches,
beginning with Illumina's ``sequencing by synthesis,'' which allows vastly 
greater throughput due to massive parallelism, low cost, and simple preparation
of samples.
Importantly to this Review, Illumina sequencing, as well as other approaches
such as SOLiD, Ion Torrent, and 454 pyrosequencing, does not read a single DNA
molecule end-to-end as one could a file from a hard drive.
Instead, DNA molecules are chopped into many small fragments, or 
\emph{reads} (Figure~\ref{fig:shotgun}), 
which are read concurrently, and which must be reassembled in the correct order 
if an entire genome is to be read.
Current reads most typically range from 50 to 200 bases long, though longer reads are
available with some technologies (e.g., PacBio).
This so-called ``shotgun sequencing'' is analogous to the problem of 
reconstructing a large file from small, 
randomly-ordered reads, or a shredded paper document.
However, these NGS reads overlap, and they are many in number, so 
reconstructing an entire genome is feasible.
No sequencing technology is completely infallible, so sequencing machines also
provide a \emph{quality score} (or measure of the confidence in the DNA base
called) associated with each position.
Thus, an NGS read is a string of DNA letters, coupled with a string of ASCII
characters that encode the quality of the base call.
A sequencing run will produce many of these reads, typically stored in a FASTQ
file.

Another important type of biological data is gene expression, or the quantity of
a particular gene product (usually a protein) present in the cell.
Gene expression data is useful for relating genotype to phenotype, and is 
important in studying diseases including cancer.
Usually obtained through microarrays or RNA-seq technologies, expression data is
quantitative, as each gene from a sample is associated with a numeric expression
level, and high-dimensional, as many thousands of genes may be analyzed at a
time.

@@

Chemical compounds represent another source of structural data, more general
than proteins.
Determined by X-ray crystallography, NMR, electron microscopy and other 
techniques, compounds---many of which are useful pharmaceuticals---can be
represented as labeled graphs.
In such a graph, the vertices are atoms (labeled with a chemical element) and
the edges are bonds between atoms (labeled with a bond type, such as single or
double).
Chemical compounds are stored in databases such as PubChem, and can be described
as adjacency matrices.

Networks are frequently used to represent biological data, such as the genetic
and physical interactions among proteins, such as those in metabolic 
pathways~\cite{berger2013computational}.

\section{Challenges with biological data}

Given DNA or RNA reads from NGS technologies, the first task is to 
\emph{assemble} those fragments of sequence (which have associated quality 
scores) into contiguous sequences.
The task of \emph{de novo} assembly is beyond the scope of this Review, but is
possible because reads overlap and because the sequence is covered by many such
overlapping reads~\cite{zhang2011practical}; the \emph{de Bruijn graph} is a 
commonly-used data structure.
Often, however, a \emph{reference genome} (or in the case of RNA, 
\emph{transcriptome}) is available for the organism being sequenced; the
establishment of a human reference genome was indeed the purpose of the Human
Genome Project.
When a reference sequence is available, NGS reads must be \emph{mapped} onto
this reference (Figure~\ref{fig:ngs-pipeline}).
Mapping allows the differences between the newly-sequenced genome and the reference
to be analyzed; these differences may include single-nucleotide polymorphisms
(SNPs, which are the genetic analogue to bit-flips) (Figure~\ref{fig:snps}), 
insertions or deletions, or larger-scale changes in the genome.
While reference-based read mapping is a fundamentally simpler problem than
\emph{de novo} assembly, it is still computationally complex, as gigabytes or
terabytes of reads must each be mapped onto the reference genome, which can 
range from millions (for bacteria) to billions (for mammals).

@@

Given a sequenced genome, the next natural questions are what genes (genomic
regions that code for proteins) are present, what structure each resulting
protein takes, and what biological function it performs.
Identifying likely genes is a well-studied problem~\cite{burge1998finding} 
beyond the scope of this Review.
However, determining evolutionary relationships, structure, and function are
at the heart of current research in computational biology.
Since some organisms (known as \emph{model organisms}) are better-studied than
others, and evolution is known to conserve sequence, structure, and function,
a powerful approach to determine these attributes is to search for similar
sequences about which more is known.
This so-called \emph{homology search} entails searching for \emph{approximate}
matches in databases of known gene or protein sequences.
Basic Local Alignment Search Tool (BLAST)~\cite{altschul1990basic} has been
the standard tool for performing homology (similarity) search on databases of nucleotide and
protein sequences.
BLAST takes a ``seed-and-extend'' approach; it looks for small, $k$-mer matches
that might lead to longer matches, and greedily extends them, ultimately 
producing a \emph{sequence alignment} between a query and each potential 
database hit.
However, BLAST's running time scales linearly with the size of the database being
searched, which is problematic as sequence databases continue to grow at a
faster rate than Moore's law.

Driven by the plummeting costs of next-generation sequencing, the 1000 Genomes Project~\cite{10002012integrated} is pursuing a broad catalog of human 
variation; instead of producing a single reference genome for a species, many
complete genomes are catalogued.
Likewise, WormBase~\cite{harris2014wormbase} and 
FlyBase~\cite{tweedie2009flybase} are cataloguing many different species and
strains of the \emph{Caenorhabditis} worm and \emph{Drosophila} fruit fly, 
respectively.
These genomes are enabling cross-species inference, and thus insights into 
evolution~\cite{berger2013computational}.

On a potentially even larger scale is the growth of metagenomic data.
Metagenomics is the study of the many genomes (bacterial, fungal, and even 
viral) that make up a particular environment.
Such an environment could be soil from a particular region (which can lead to 
the discovery of new 
antibiotics~\cite{gillespie2002isolation,forsberg2012shared}), or it could be 
the human gut, whose microbiome has been linked to human-health concerns 
including Autism Spectrum Disorder~\cite{macfabe2012short}, 
Crohn's Disease~\cite{gevers2014treatment,manichanh2006reduced}, and 
obesity~\cite{greenblum2012metagenomic}.
Metagenomics fundamentally asks what organisms are present, and (in the case
of a microbiome such as the gut) what metabolic functions it can accomplish as
a whole.
One way of addressing this problem is to attempt to map NGS reads from a metagenomic sample
onto a set of reference genomes that are expected to be present.
This is exactly the read-mapping problem discussed above, but with many 
reference genomes, compounding the computational requirements.
A second way is to perform homology search on a protein sequence database;
exact or nearly-exact matches imply the presence of a species, while more 
distant hits may still give clues to function.
For this task, BLASTX~\cite{altschul1990basic} is commonly
used to translate nucleotide reads into their possible protein sequences, and
search for them in a protein database.
The difficulty is that the datasets required to shine any light on these 
questions, namely from ``shotgun'' metagenomics, are gigantic and vastly more 
complex than 16S metagenomic studies that merely classify the bacteria based on 
their divergence with respect to a specific gene, resulting in identification 
problems for certain bacterial, as well as viral, species and 
genera~\cite{janda200716s}.

While gene expression data lends itself to cluster analysis and probabilistic 
approaches, the high dimensionality and noise in the data presents significant 
challenges~\cite{nguyen2002tumor}.
Principal Component Analysis has shown promise in reducing the dimensionality
of gene expression 
data~\cite{raychaudhuri2000principal,alter2000singular,yeung2001principal,misra2002interactive}.

@@

The study of drugs and their targets based on chemical structure and function
is known as chemogenomics~\cite{bredel2004chemogenomics}.
In the fields of drug discovery and drug repurposing, the prediction 
of biologically active compounds is an important task. 
Computational high-throughput screening eliminates many compounds from 
laborious wet-lab consideration, but even computational screening can be 
time-consuming.
Notably, large chemical compound databases, such as PubCHEM,
cannnot be searched on a laptop computer with standard tools (e.g., SMSD).

With biological networks, challenges lie in integrating such data in order to 
deduce gene function and disease pathways.
Several approaches have taken advantage of the particular topologies of 
biological
networks, such as diffusion-based approaches, which explore the topology of
networks through random walks.
In protein interaction networks, some nodes are high-degree \emph{hubs,} which 
interact with
many partners; these interactions do not imply shared function as strongly as
non-hub interactions; such information needs to be teased apart.
As new functional annotations and protein-protein interactions are discovered,
the number and size of these networks grows, making related problems more 
intractable~\cite{berger2013computational}.

@@

\section{State-of-the-art approaches to meet these challenges}

Techniques for reference-based read mapping typically rely on algorithmic 
approaches such as the Burrows-Wheeler transform (BWT)~\cite{burrows1994block}, 
which provides efficient string 
compression through a reversible transformation, while the FM-index data 
structure~\cite{ferragina2000opportunistic} is a compressed substring index, 
based on the BWT, which provides efficient storage as well as fast search.
BWA (Burrows-Wheeler Aligner)~\cite{li2009fast} uses the BWT, while the 
Bowtie2~\cite{langmead2012fast} aligner further relies on the FM-index for 
efficient mapping of NGS reads.
The Genome Multitool (GEM) mapper~\cite{marco2012gem} also uses an FM-index 
coupled with dynamic programming in a compressed representation, in order to 
prune the search space when mapping reads to a reference genome.
Masai~\cite{siragusa2013fast} uses an ``approximate seed'' approach to indexing
the space of possible matches, likewise pruning the search space.
Even with these approaches, read mapping remains a significant bottleneck in
genomic research~\cite{hach2010mrsfast}.

Recent advances in metagenomic search tools have relied on two advances over
BLASTX: indexing and alphabet reduction.
RapSearch2~\cite{zhao2012rapsearch2} relies on alphabet reduction and a 
collision-free hash table.
The alphabet reduction, as it is reversible, can be thought of as a form of
lossless compression; a 20-letter amino acid alphabet is mapped onto a smaller
alphabet, with offsets stored to recover the original sequence in the full
alphabet.
The hash-table provides an efficient index of the database to be searched.
DIAMOND~\cite{buchfink2014fast} also relies on alphabet reduction, but uses
``shaped seeds'' instead of simple $k$-mer seeds to index the database.
DIAMOND demonstrates search performance three to four orders of magnitude faster
than BLASTX, but it is still linear in the size of the database being searched.

Recent work on gene expession has explored additional ways to exploit the high-dimensional 
structure of the data.
SPARCLE (SPArse ReCovery of Linear combinations of Expression)~\cite{prat2011recovering}
brings ideas from compressed sensing~\cite{candes2005decoding,candes2006stable,donoho2006compressed} 
to gene expression analysis.
Specifically, SPARCLE relies on the idea that a randomly-chosen set of points 
are unlikely to be close in a lower-dimensional linear subspace.
Thus, SPARCLE approximates a gene's expression profile as a linear combination
of a few other genes' profiles, and then applies supervised learning to
predict relationships between genes.
Another recent and novel approach to exploiting the structure of gene expression
space is Parti (Pareto task inference)~\cite{hart2015inferring}, which describes a set of
data as a polytope, and infers the specific tasks represented by vertices of
that polytope from the features most highly enriched at those vertices.
In human breast tumors, and in mouse tissues, the expression data were
well-described by tetrahedra whose vertices were enriched for different tumor
types and biological functions, from which~\cite{hart2015inferring} infer four distinct
tasks.

An area of great interest in integrating network data is in global multiple network 
alignment, which is based on network topology and conserved biological function (IsoRank~\cite{singh2008global} and IsoRankN~\cite{liao2009isorankn}).
IsoRank's approach is similar to the PageRank algorithm~\cite{page1999pagerank}, while
IsoRankN produces a \emph{Personalized PageRank} alignment, similar to the
PageRank-Nibble algorithm~\cite{andersen2006local}.
Diffusion state distance (DSD)~\cite{cao2013going} is a diffusion-based approach
which defines a distance metric based on random walks.
DSD penalizes high-degree nodes, and outperforms other graph-theoretic distance
functions when applied to standard approaches such as neighbor voting for function annotation.
Diffusion component analysis (DCA)~\cite{cho2015diffusion} amounts to a network-based 
analogue of principle component analysis; DCA reduces the dimensionality of a
network in such a way that function annotation can be performed in the reduced 
space, significantly reducing running time while still providing competitive
prediction accuracy.

@@

One key insight as to why the ``compressive omics'' approach allows computation
that scales sublinearly in the size of the data is that such data sets exhibit
low \emph{metric entropy}~\cite{yu2015entropy}.
That is, for a given cluster radius $r_c$ and a database $D$, the number $k$ 
of clusters needed to cover $D$ is bounded by $N_{r_c} (D)$, the metric 
entropy, which is relatively small compared to
$|D|$, the number of entries in the database.
In contrast, if the points were uniformly distributed about the Cartesian space,
$N_{r_c} (D)$ would be larger.

A second key insight is that the biological data sets considered 
by~\cite{yu2015entropy} exhibit low \emph{fractal dimension}.
That is, within some range of radii $r_1$ and $r_2$ about an arbitrary point
in the database $D$, the fractal dimension $d$ is
$d=\frac{\log (n_2 / n_1)}{ \log (r_2 / r_1)}$, where $n_1$ and $n_2$ are the
number of points within $r_1$ and $r_2$ respectively.

Cluster-based search, as exemplified by ``compressive omics,'' can perform
approximate search within a radius $r$ of a query $q$ on a database $D$ with 
fractal dimension $d$ and metric entropy $k$ at the scale $r_c$ in time
proportional to
\begin{gather*}
    O\Bigg(
    \underbrace{k}_{\textrm{metric entropy}} +
    \overbrace{\left|B_D(q,r)\right|}^{\textrm{output size}}
    \underbrace{\left(\frac{r+2r_c}{r}\right)^d}_{\textrm{scaling factor}}
     \Bigg) .
\end{gather*}

Given this formalization, the ratio $\frac{|D|}{k}$ provides an estimate of the 
speed-up factor for the coarse search component compared to a full linear 
search.
The time-complexity of the fine search is exponential in the fractal dimension
$d$, which can be estimated globally by sampling the local fractal dimension 
over a dataset.
Table~\ref{fractal} provides the fractal dimension $d$ sampled at typical query
radii, as well as the ratio $\frac{|D|}{k}$, for nucleotide sequence, protein 
sequence, protein structure, and chemical compound databases.

Biological data sets exhibit redundancy, and are constrained to subspaces by
physical laws.
This combination results in low fractal dimension and low metric entropy 
relative to the size of the data set, which suggests that ``compressive omics''
will provide the ability for computation to scale with massively growing data.

\begin{table}
\label{fractal}
\caption{Metric entropy ratio (ratio of clusters to entries in database) and
fractal dimension at typical search radii for four biological data sets.
NCBI's ``NR'' non-redundant protein sequence and ``NT'' non-redundant nucleotide sequence databases are from June, 2015. Protein Databank (PDB) is from July, 2015. PubChem is from October, 2013.}
\begin{tabular}{ccc}
\hline
Data set & Metric entropy ratio & Fractal dimension \\        
\hline
Nucleotide sequences (NCBI NT) & 7:1 & 1.5\\
\hline
Protein sequences (NCBI NR) & 5:1 & 1.6\\
\hline
Protein structure (PDB) & 10:1 & 2.5\\
\hline
Chemical structure (PubChem) & 11:1 & 0.2\\
\hline
\end{tabular}
\end{table}
